Java supports three jump statements {\bf break, continue}, and {\bf return}.
\vskip 1cm
{\bf break}
\vskip 3mm
The {\bf break} statement has three uses. First, it terminates a statement sequence in a {\bf switch} statement. Second, it can be used to exit a loop. Third, it can be used as a structured form of {\bf goto}.
\vskip 1mm
By using {\bf break}, you can force immediate termination of a loop, bypassing the conditional expression and any remaining code in the body of the loop. When a {\bf break} statement is encountered inside a loop, the loop is terminated and program control resumes at the next statement.

\vskip 1mm
When used inside a set of nested loops, the {\bf break} statement will only break out of the innermost loop.

\vskip 1mm
Here are two important points to remember about {\bf break}. First, more than one {\bf break} statement may appear in a loop. Be careful. Too many {\bf break} statements can destructure your code. Second, the {\bf break} that terminates a {\bf switch} statement affects only that {\bf switch} statement and not any enclosing loops.

\vskip 1in
\filbreak
{\bf Using break as a Form of Goto}
\vskip 3mm
The {\bf break} statement can also be used as a structured form of the {\bf goto} statement. Java does not have a goto statement because it provides a way to branch in an arbitrary and unstructured way. There are places where the goto is a valuable and legitimate construcut for flow control. The goto can be useful when you are exiting from a deeply nested set of loops. To handle such situations, Java defines an expanded form of the {\bf break} statement. By using {\bf break} you can break out of one or more blocks. These blocks don't need to be part of a loop or a {\bf switch}. They can be any block. You can specify precisely where execution will resume, because this form of {\bf break} works with a label.

\vskip 1mm
Here is the general form of the labeled {\bf break} statement:

$$\hbox{\tt break {\it label};}$$

\vskip 1mm
Most often, {\it label} is the name of a label that identifies a block of code. Than be a standalone block of code but it can also be a block that is the target of another statement, When this form of {\bf break} executes, control is transferred out of the named block. The labeled block must enclose the {\bf break} statement, but it does not need to be the immediately enclosing block. This means, for example, that you can use a labeled {\bf break} statement to exit from a set of nested blocks. But you cannot use {\bf break} to transfer control out of a block that does not enclose the {\bf break} statement.

\vskip 1mm
To name a block, put a label at the start of it. A {\it label} is any valid Java identifier followed by a colon. Once you have labeled a block, you can then use this label as the target of a {\bf break} statement. Doing so causes execution to resume at the {\it end} of the labeled block.

\vskip 1mm
One of the most common use for a labeled break statement is to exit from nested loop. When the inner loop breaks to the outer loop, both loops have been terminated. Keep in mind that you cannot break to any label which is not defined for an enclosing block.

\vskip 1in
\filbreak
{\bf Using Continue}
\vskip 3mm
You might want to continue running the loop but stop processing the remainder of the code in its body for this particular iteration. The {\bf continue} statement performs such an action. In {\bf while} and {\bf do-while} loops, a {\bf continue} statement causes control to be transferred directly to the conditional expression that controls the loop. In a {\bf for} loop, control goes first to the iteration portion of the {\bf for} statement and then to the conditional expression. For all three loops, any intermediate code is bypassed.

\vskip 1mm
As with the {\bf break} statement, {\bf continue} may specify a label to describe which enclosing loop to continue.

\vskip 1mm
Good uses of {\bf continue} are rare. For those special circumstances in which early iteration is needed, the {\bf continue} statement provides a structured way to accomplish it.

\vskip 1in
\filbreak
{\bf return}

\vskip 3mm
The {\bf return} statement is used to explicitly return from a method. That is, it causes program control to transfer back to the caller of the method.

\vskip 1mm
At any time in a method, the {\bf return} statement immediately terminates the method in which it is executed.

\vskip 1mm
As soon as {\bf return} is executed, control passes back to the caller.

\filbreak
\vfill\eject
\bye
