{\bf Selection}
\vskip 3mm

{\bf General form of {\tt if} Statements}
\vskip 1mm
$$\vbox{\+\tt if \cleartabs& ({\bf condition}) $\{$\cr
	\+&\bf statement1; \cr
	\+$\}$\cr
	\+\tt else $\{$\cr
	\+&\bf statement2;\cr
	\+$\}$\cr}$$

\vskip 1mm
Each statement may be a single statement or a compound statement enclosed in curly braces. The condition is any expression that returns a {\bf boolean} value.

\vskip 1cm
\filbreak
\vskip 1mm
{\bf Nested {if} Statement}
$$\vbox{\+\tt if \cleartabs& ({\bf condition}) $\{$\cr
	\+&\tt if \cleartabs& ({\bf condition}) $\{$\cr
	\+&&\bf statement;\cr
	\+& $\}$ \cr
	\+&\tt else $\{$ \cr
	\+&&\bf statements; \cr
	\+& $\}$ \cr
	\+\tt else $\{$\cr
	\+\bf statement;\cr
	\+$\}$ \cr}$$

\vskip 1mm
An {\tt else} statement always refers to the nearest {\tt if} statement.

\vskip 1cm
\filbreak
\vskip 1mm
{\bf General form of {\tt if-else-if} Statements}
\vskip 1mm
$$\vbox{\+\tt if \cleartabs& ({\bf condition}) $\{$\cr
	\+&\bf statement;\cr
	\+\tt else if ({\bf condition}) $\{$\cr
	\+&\bf statement;\cr
	\+\tt else if ({\bf condition}) $\{$\cr
	\+$\}$\cr
	\+&\bf statement;\cr
	\+$\}$\cr
	\+$\vdots$\cr
	\+\tt else $\{$\cr
	\+&\bf statement;\cr
	\+\tt $\}$\cr}$$

\vskip 1mm
The {\tt if} statements are executed from the top down. As soon as one of the conditions controlling the {\tt if} is {\bf true}, the statement associated with that {\tt if} is executed, and the rest of the code is bypassed. If none of the conditions is true, then the final {\tt else} statement will be executed. The final {\tt else} statement is the default condition; that is, if all other conditional test fail, then the last {\tt else} statement is performed. If there is no final {\tt else} and all conditions are {\bf false}, then no action will take place.

\vskip 1cm
\filbreak
{\bf General form of {\tt switch} statement}
\vskip 1mm

$$\vbox{\+\tt switch \cleartabs& ({\bf expression}) $\{$\cr
	\+\tt case {\bf value1}:\cr
	\+&\bf statement;\cr
	\+&\bf break;\cr
	\+\tt case {\bf value2}: \cr
	\+$\vdots$\cr
	\+\tt case {\bf valueN}:\cr
	\+&\bf statement;\cr
	\+$\}$\cr}$$

\vskip 1mm
The expression must be of type {\tt byte, short, int, char, String}, or an enumeration. Each value specified in the {\tt case} statements must be a unique constant expression. Duplicate {\tt case} values are not allowed. The type of each value must be compatible with the tyoe of expression.
\vskip 1mm
A constant expression is an expression that yields a primitive type or a String, and whose value can be evaluated at compile time to a literal.

\vskip 1mm
The {\tt break} statement is used inside the {\tt switch} statement to terminate a statement sequence. When a {\tt break} statement is encountered, execution branches to the first line of code that follows the entire switch statement.

\vskip 1mm
The {\tt break} statement is optional. If you omit the {\tt break}, execution will continue on into the next {\tt case}.

\vskip 1cm
\filbreak
{\bf Important features of the {\tt switch} statement}
\vskip 1mm
\qquad$\bullet$ The {\tt switch} differes from the {\tt if} in the fact that {\tt switch} can only test for equality, whereas {\tt if}
\vskip 1mm
\qquad\quad can evaluate any type of Boolean expression. The {\tt switch} looks only for a match betwen the
\vskip 1mm
\qquad\quad value of the expression and one of its {\tt case} constants.

\vskip 2mm
\qquad$\bullet$ No two {\tt case} constants in the same {\tt switch} can have identical values.

\vskip 2mm
\qquad$\bullet$ A {\tt switch} statement is usually more efficient than a {\it set} of nested {\tt if}s.
\filbreak
\vfill\eject
\bye
