{\bf Bash Reference Manual Definitions}
\vskip 1mm
\hrule

\vskip 1cm
{\tt POSIX}
\vskip 1mm
A family of open system standards based on Unix. Bash is primarily converned with the Shell and Utilities portion of the {\tt POSIX 1003.1} standard.

\filbreak
\vskip 3mm
{\tt blank}
\vskip 1mm
A space or tab character.

\filbreak
\vskip 3mm
{\tt builtin}
\vskip 1mm
A command that is implemented internally by the shell itself, rather than by an executable program somewhere in the file system.

\filbreak
\vskip 3mm
{\tt control operator}
\vskip 1mm
A {\tt token} that performs a control function. It is a {\tt newline} or one of the following: '$||$', '$\&\&$', '$\&$', ';', ';;', ';$\&$', ';;$\&$', '$|$', '$|\&$', '(', or ')'.

\filbreak
\vskip 3mm
{\tt exit status}
\vskip 1mm
The value returned by a command to its caller. The value is restricted to eight bits, so the maximum value is 255.

\filbreak
\vskip 3mm
{\tt field}
\vskip 1mm
A unit of text that is the result of one of the shell expansions. After expansion, when executing a command, the resulting fields are used as the command name and arguments.

\filbreak
\vskip 3mm
{\tt filename}
\vskip 1mm
A string of characters used to identify a file.

\filbreak
\vskip 3mm
{\tt job}
\vskip 1mm
A set of processes comprising a pipeline, and any processes descended from it, that are all in the same process group.

\filbreak
\vskip 3mm
{\tt job control}
\vskip 1mm
A mechansim by which users can selectively stop (suspend) and restart (resume) execution of processes.

\filbreak
\vskip 3mm
{\tt metacharacter}
\vskip 1mm
A character that, when unquoted, separates words. A metacharacter is a {\tt space}, {\tt tab}, {\tt newline}, or one of the following characters: '$|$, '$\&$', ';', '(', ')', '$<$, or '$>$.

\filbreak
\vskip 3mm
{\tt name}
\vskip 1mm
A {\tt word} consisting solely of letters, numbers, and underscores, and beginning with a letter or underscore. {\tt Name}s are used as shell variable and function names. Also referred to as an {\tt identifier}.

\filbreak
\vskip 3mm
{\tt operator}
\vskip 1mm
A {\tt control operator} or a {\tt redirection operator}. Operators contain at least one unquoted {\tt metacharacter}.

\filbreak
\vskip 3mm
{\tt process group}
\vskip 1mm
A collection of related processes each having the same process group ID.

\filbreak
\vskip 3mm
{\tt process group ID}
\vskip 1mm
A unique identifier that represents a {\tt process group} during its lifetime.

\filbreak
\vskip 3mm
{\tt reserve word}
\vskip 1mm
A {\tt word} that has a special meaning to the shell. Most reserved words introduce shell flow control constructs, such as {\tt for} and {\tt while}.

\filbreak
\vskip 3mm
{\tt return status}
\vskip 1mm
A synonym for {\tt exit status}

\filbreak
\vskip 3mm
{\tt signal}
\vskip 1mm
A mechanism by which a process may be notified by the kernel of an event occuring in the system.

\filbreak
\vskip 3mm
{\tt special builtin}
\vskip 1mm
A shell builtin command that has been classified as special by the {\tt POSIX} standard.

\filbreak
\vskip 3mm
{\tt token}
\vskip 1mm
A sequence of characters considered a single unit by the shell. It is either a {\tt word} or an {\tt operator}.

\filbreak
\vskip 3mm
{\tt word}
\vskip 1mm
A sequence of characters as a unit by the shell. Words may not include unquoted {\tt metacharacters}.
%$$\vbox{\+\bf \cleartabs& \cr
%	\+\cr
%	\+\cr
%	\+\cr}$$

\vfill\eject
\bye
